\documentclass[12pt,a4paper,article]{memoir}
%
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage[dvipsnames]{xcolor}
\usepackage{mathpazo}
\usepackage[T1]{fontenc}
\linespread{1.05}
\date{}
\title{Oligo-poly\\
\vspace{10pt}
\large
A library for working with polynomial rings under oligomorphic group actions.
$ $\\}
%
\author{Arka Ghosh}
%
\newcommand{\struct}{\mathcal{X}}
\newcommand{\dloQ}{\mathcal{Q}}
\newcommand{\ordZ}{\mathcal{Z}}

\newcommand{\N}{\mathbb{N}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\F}{\mathbb{F}}

\newcommand{\aut}[1]{\mathsf{Aut}\left(#1\right)}
\newcommand{\poly}[2]{#1[#2]}
\newcommand{\setof}[2]{\left\{#1\ \mid\ #2 \right\}}
\newcommand{\idGen}[1]{\langle#1\rangle}
\newcommand{\eqIdGen}[1]{\langle#1\rangle_{\textnormal{eqv}}}

\newcommand{\arka}[1]{}
%{\textcolor{red}{Arka:#1}}
\newcommand{\emphdef}[1]{\emph{\textcolor{BrickRed}{#1}}}
\newcommand{\code}[1]{\texttt{\textcolor{magenta}{#1}}}

\newtheorem{theorem}{Theorem}

\theoremstyle{definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{todo}[theorem]{Todo}
%
\begin{document}
\maketitle
\raggedbottom
%
\tableofcontents
%
\pagebreak
\noindent
The reader is suggested to look at the most recent version of this document which can be found in the url below.
\[
\text{
\url{https://github.com/arkagh0sh/hilbert}
}
\]
The purpose of this library is to implement (in Rust) the equivariant Buchberger algorithm given in \cite{GL25}.
This document is mostly self-complete in terms of definitions.
%
\paragraph*{Structure of this document}
%
We start with explaining the theoretical background in \Cref{sec:theory}.
%\Cref{sec:how} gives basic instructions for using this library.
\Cref{sec:impl} describes the important aspects of our implementation and gives basic instructions for usage.
%
\chapter{Theoretical background}\label{sec:theory}
%

\section{Relational structures}
%
A \emphdef{relational structure} $\struct = (X,R_1,\dots,R_n)$ consists of a
set $X$ (called the \emphdef{domain} of the relational structure) and relations $R_i$ defined on the set $X$.
%
\begin{example}
Let $Q$ be the set of rational numbers.
We can consider the structure $\dloQ = (Q,<)$ where is $<$ the usual ordering.
\end{example}
%
\begin{example}
A graph $\mathcal{G} = (V,E)$ is another example of a relational structure,
with the set $V$ of vertices being the domain and the edge relation $E$ being the only relation.
\end{example}

By abuse of notation we sometime use $\struct$ to denote the set underlying $X$.
For instance,
we write $x \in\struct^n$ to mean that $x$ is an element of $X^n$.

\section{Automorphism groups}

An \emphdef{automorphism} of the relational structure $\struct$ is a bijection from $\struct$ to itself that preserves and reflects the structure.
%
\begin{example}
An automorphism of $\dloQ$ is an order preserving bijection from $\mathcal{Q}$ to itself.
\end{example}
%
\begin{example}
If $\struct$ is a graph,
then an automorphism of $\struct$ is an isomorphism of the graph with itself.
\end{example}
%
The set of automorphisms of a relational structure $\struct$ is denoted as \emphdef{$\mathbf{Aut}(\struct)$}.
%
\section{Oligomorphic structures}
%
A relational structure $\struct^n$ is called \emphdef{oligomorphic} if for every $n\in\N$,
the set $\struct^n$ splits into finite many orbits.
%
\begin{example}
The structure $\dloQ$ is oligomorphic.
For illustration, for $n = 2$, $\dloQ^2$ splits into three orbits
\[
\setof{(a,b)\in\dloQ^2}{a < b},
\setof{(a,b)\in\dloQ^2}{a = b}
\text{ and }
\setof{(a,b)\in\dloQ^2}{a > b} \ .
\]
\end{example}
%
\begin{example}
The structure $\ordZ = (Z,<)$ of integers with order is not oligomorphic.
For instance,
the set $\ordZ^2$ splits into infinite many orbits of the form
\[
\setof{(a,b)}{a - b = d}
\]
for $d\in Z$.
\end{example}
%
\section{Polynomial ideals}
%
We use $\poly{\F}{X}$ to denote the polynomial rings with coefficients from the field $\F$ and variables from the set $X$.
A subset $J$ of polynomials is called an \emphdef{ideal} if
\begin{enumerate}
\item the zero polynomial $0$ is inside $J$,
\item the sum $f + g$ of any two polynomials $f$ and $g$ inside $J$ is also inside $J$, and
\item the product $h\cdot f$ of any polynomial $h$ in the ring $\poly{\F}{X}$ with any polynomial $f\in J$ is also in $J$.
\end{enumerate}
%
\begin{notation}
For a function $\alpha : X\to \F$ and $f\in\poly{\F}{X}$ let $f(\alpha)$ denote the instantiation of $f$ at $\alpha$.
For instance, if $f = x^2 - y$, $\alpha(x) = 1$ and $\alpha(y) = 1.5$,
then $f = -0.5$.
\end{notation}
%
\begin{example}
For any subset $S\subseteq X\to \F$,
the set
\[
\setof{f\in\poly{\F}{X}}{f(\alpha) = 0\text{ for all }\alpha\in S}
\]
is an ideal inside $\poly{\F}{X}$.
\end{example}
%
\section{Equivariant ideals}
%
For a relational structure $\struct$,
we use $\mathbb{Q}[\struct]$ to denote the polynomial ring whose variables are elements of $\struct$.

Instead of using $\struct$ as the set of variables some articles use variables $a_{x}$ indexed by elements $x\in\struct$.
But using $\struct$ itself makes the notation simpler.

The group $\mathbf{Aut}(\struct)$ of automorphisms of $\struct$ acts on $\mathbb{Q}[\struct]$ in a variable-wise manner.
For instance, for $\pi\in\mathbf{Aut}(\struct)$ and $x,y\in\struct$
$$\pi(x^2 - 2xy^2 + 3) =
\pi(x)^2 - 2\pi(x)\pi(y)^2 + 3\ .$$

An ideal in $\poly{\Q}{\struct}$ is called \emphdef{equivariant},
if it is invariant under the action of $\aut{\struct}$.
%
\begin{example}
The set $J_0$ of polynomials in $\poly{\Q}{\struct}$ whose coefficients add up to zero is an equivariant ideal.

Equivalently, $J_0 = \setof{f\in\poly{\Q}{\struct}}{f(\mathbf{1}) = 0}$ where $\mathbf{1}$ is the function $\dloQ\to\Q$ defined as $\mathbf{1}(a) = 1$ for all $a\in\dloQ$.
\end{example}
%
For a subset $G\subseteq\poly{\F}{\struct}$,
the \emphdef{ideal $\idGen{G}$ generated by $G$} is the smallest ideal containing $G$.
For instance, $\idGen{\setof{a - 1}{a\in\dloQ}} = J_0$

Similarly, for a subset $G\subseteq\poly{\F}{\struct}$,
the \emphdef{equivariant ideal $\eqIdGen{G}$ generated by $G$} is the smallest equivariant ideal containing $G$.
For instance $\eqIdGen{\{a - 1\}} = J_0$ for any $a\in\dloQ$.
%
An (equivariant) ideal is called \emphdef{finitely generated} if it is generated by a finite set.

Note that equivariant makes finite generation easier.
For instance, one can prove that $J_0$ is not finite generated as an ideal,
but since $J_0 = \eqIdGen{\{a - 1\}}$,
it is finitely generated as an equivariant ideal.
%
\section{Equivariant Gröbner bases}
%
We say $\struct$ has the \emphdef{Hilbert basis property} if for every field $\F$, every equivariant ideal $J$ in $\poly{\F}{\struct}$ is generated by a finite set.

Theorems 11 and 12, and Corollary 13 of \cite{GL24} characterised the Hilbert's basis property of relational structures.

\arka{TODO:expand starting from this part}

We call relational $\struct$ to be \emphdef{ordered} if one of the relations of  $\struct$ induces a linear order on $\struct$.
If we restrict our attention to ordered relational structure,
then the results of \cite{GL24} collapse to the following theorem.
%
\begin{theorem}
A ordered relational structure satisfies the Hilbert's basis property if and only if finite $\N$-labelled substructures of $\struct$ are well-quasi-ordered under embeddability.
\end{theorem}
%
The article \cite{GL25} builds on their results by defining Gr\"{o}bner bases of equivariant ideals and observes that if a relational structure $\struct$ satisfies the Hilbert's basis property,
then every equivariant ideals in $\poly{\F}{\struct}$ have finite Gr\"{o}bner bases.
The central result result of \cite{GL25} is to give a modified version of the equivariant extension of the Buchberger's algorithm that takes as input a finite set $H$ of polynomials in $\poly{\Q}{\struct}$ and outputs a Gr\"{o}bner basis of the equivariant ideal generated by $H$.
They show that this algorithm terminates when $\struct$ is a ordered relational structure satisfying the Hilbert's basis property.
%
%%
%This chapter is divided into two sections.
%The first section describes how to define and work with relational structures.
%%
%\section{Relational structures}\label{sec:how str}
%
%\subsection{Pre-defined relational structures}\label{sec:pre def str}
%
%\subsubsection{Dense linear order without endpoints}\label{sec:how dlo}
%
%\subsection{Defining new relational structures}\label{sec:how new str}
%
%
%\section{Polynomial rings}
%
%\subsection{Ideals}
%
%\subsection{Buchberger's algorithm}
%
%\subsection{Equivariant Buchberger's algorithm}
%
\chapter{Implementation details}\label{sec:impl}
%
The implementation is divided into three parts.
The first part gives a \emph{framework} to write equivariant algorithms.
The second part is a simple implementation of polynomial rings $\poly{\Q}{\struct}$ where $\struct$ can be any arbitrary data-type.
The third part combines these two and implements the equivariant Buchberger's algorithm.

The work regarding the third is still in progress,
so we omit the details of this part.
%
\section{Framework for equivariant algorithms}
%
We are heavily influenced by \code{LOIS} and \code{N$\lambda$}.
The main difference is that both \code{LOIS} and \code{N$\lambda$} guarantees that any program written in them is equivariant by construction.
We only guarantee that any program written following our framework is equivariant.
The benefit of doing this is that the codebase is significantly smaller and simpler.
\begin{todo}
Implement a checker that checks whether a piece of code follows the framework.
\end{todo}
%
Like \code{LOIS} and \code{N$\lambda$} we categories objects into two categories,
\emphdef{elements of orbits} (encoded by the \code{trait} \code{ElementsWithAtoms}) and \emphdef{orbits} (encoded by the \code{struct} \code{Orbit}).
\footnote{Obviously, orbits themselves can be elements of orbits.
However, since we only deal with equivariant orbits as of now,
orbits of orbits are singleton.
\arka{This is schizophrenic}}
%
\subsection{\code{trait AtomsWithOrd}}
%
This gives the framework for working with \emph{computationally oligomorphic relational structures}.
To make a \code{struct A} satisfy this \code{trait},
one has to implement two basic operations.
\begin{enumerate}
\item \code{in\_same\_orbit}: This function checks if two \code{BTreeSet<A>} are in the same orbit.
\item \code{orbits\_in\_product} : This functions takes two orbits $B$ and $C$ of \code{BTreeSet<A>} as input and outputs the orbits of \code{(BTreeSet<A>,BTreeSet<A>)} in $B\times C$. 
\end{enumerate}
%
\subsubsection{Defining relational structures}
%
One must be careful while defining a new \code{struct} as a relational structure.
Our recommended way of defining a relational structure is to use a wrapper class so the actual value can be accessed by the functions that are defined in the struct.

For illustration we define \code{DLO} as
\color{magenta}
\begin{verbatim}
#[derive(PartialEq, PartialOrd,Eq,Ord)]
pub struct DLO {
    val : Rational
}
\end{verbatim}
\color{black}
%
Observe that the field \code{val} is private.
Therefore one cannot write codes that uses operations other than the order and equality on \code{DLO}.
For instance, this disallows code snippets such as
\color{magenta}
\begin{verbatim}
let r : Rational = (x.val + y.val)/2;
let z : DLO = DLO::new(r);
\end{verbatim}
\color{black}
%
\arka{one can still write a piece of code that runs on a while loop and computes an element greater than a given element}
%
\subsection{\code{trait ElementsWithAtoms<A>}}
%
This is the \code{trait} for elements of orbits constructed from the atoms \code{A}.
To implement this \code{trait} for a \code{struct B} one has to implement three functions.
\begin{enumerate}
\item \code{support()}: This computes the support of an element.
\item \code{project\_to\_support(X : Orbit<A,B>, S : BTreeSet<A>)}: This takes an orbit \code{X} inside \code{B} finite subset of elements of \code{A} as inputs and outputs all elements in \code{X} that are supported by \code{S}.
\end{enumerate}
%
\subsection{\code{struct Orbit<A,B>}}
%
This is the structure of orbits of elements of another \code{struct B} under the action of the automorphism group of another \code{struct A}.
An object is given by its representative \code{rep}.
This field is kept private.
This is necessary to ensure that any function that takes orbits as input is equivariant.
%
\arka{we have to add definition of equivariance and explain it}
%
An object of the \code{struct} \code{orbit} is given by its representative \code{rep} which must satisfy the \code{trait} \code{ElementsWithAtoms}.
The \code{rep} field is hidden,
this ensures that orbits can only be accessed in specific ways (explained below) that preserves equivariance.

Orbits are considered as lists indexed by elements of
\code{BTreeSet<A : AtomsWithOrd>}.
This is consistent with the theoretical result that every orbit of set with atoms $\struct$ is isomorphic to an equivariant subset of $\binom{\struct}{n}$ for some $n\in\N$,
Orbits can be accessed in the following ways:
\begin{enumerate}
\item Given two orbits $B$ and $C$, one can check if they are equal.
\item Given an orbit $B$ and \code{S : BTreeSet<A>} one can compute the elements in the orbit supported by \code{S}.
\item Given two orbits $B$ and $C$, one can compute the list of orbits in the product $B\times C$.
\item Given an equivariant function $f$ with domain containing $B$,
one can compute $f(B)$ by applying \code{map}.
\item Given an equivariant function $f$ with domain containing $B$ and boolean output,
one can filter $B$ using $f$.
Observe that since $f$ is equivariant, this means the result of the filtering is either $B$ or $\emptyset$.
\end{enumerate}
The reason behind doing the above is the following.
First, we ensure that any function with input and output types satisfying the trait \code{ElementsWithAtoms} is equivariant.
Second, we ensure that orbits can only be accessed in specific ways that preserves equivariance.
As a consequence, any function that takes orbits as input must be equivariant.
%
\section{Polynomial rings}
%
The file \code{mon.rs} implements polynomials and basic operations on them.
The standard way to define a polynomial is to first define the variables.
For instance, the code below show how to define a variable with type \code{String},
\[
\text{\code{let x : Polynomial<F2,String> = Polynomial::var("x".to\_string());}}
\]
\arka{this needs to clearer and more detailed}
Here $F2$ is the finite field with two elements.
Variables themselves are considered as polynomials.
Once we have defined variables $y$ and $z$ in a similar manner,
the polynomial $f = x + yz$ can be defined as
\[
\text{\code{let f = x.clone() + y.clone()*z.clone()}};
\] 
The \code{.clone()} is needed to ensure that the variables can be reused.
All the usual operations (addition, subtraction, multiplication, powering) of polynomials are defined in \code{mon.rs}.

The only important thing to note here is that monomials are compared using the reverse lexicographic ordering.
%
\subsection{Polynomial ideals}
%
Polynomials ideals are implemented in \code{ideal.rs}.
An ideal is defined as a \code{struct} containing a set of its generators.
At this point the only functions implemented is \code{buchberger},
which as the name suggests,
implements the Buchberger's algorithm.

The next steps would be implement ideal membership, equality checking of ideals,
computation of sum, product and quotients.
%
\begin{thebibliography}{9}
\bibitem{GL24}
Arka Ghosh and S\l{}awomir Lasota. Equivariant ideals of polynomials, LICS 2024.
%
\bibitem{GL25}
Arka Ghosh and Aliaume Lopez. Computability of equivariant
Gr\"{o}bner bases, 2025.
%
\end{thebibliography}
%
\end{document}
